Sustainable Pathways for Irish homes in changing TIMES

### 2021-07-08 manuscript.tex
8 July 2021

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{pgf-pie}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{anyfontsize}
\usepackage{pgfplots}
\usepackage{amsmath}
\usepackage{comment}
\usepackage{booktabs}
\usepackage{rotating}
\usepackage{siunitx}
\usepackage{eurosym}
%\usepackage{multirow}

@@

Energy System Optimisation Models (ESOMs) can be used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM covers all energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.

\par For example, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK using UK TIMES Model (UKTM). As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. Li set out to incorporate a Discrete Choice Model (DCM) for TIMES to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices. While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights through more surveys and analysis.

@@

Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark. The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland, and Germany to cover import/export. Exogenous user constraints applied to the model, based on the current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030, and individual oil boilers are phased out by 2030. The results find that when investments in individual heat pumps are allowed, air-to-water heat pumps are installed in non-district heating areas. This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and reducing system costs, fuel consumption, and $CO_2$ emissions.
Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a reduced need for peak/reserve capacity investments. The Balmorel model has provided insights, but it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies. Nevertheless, the insights are valuable and the results can be soft-linked to other models for deeper analysis.

Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at the household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island. 
Two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach. Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be stored in communal batteries. When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery. While the battery location was explored, the PV system location was not, further investigation into the optimal PV system location could provide further insights. The household occupancy behaviour with and without household batteries on the PV system could alter the optimal pathway.

@@

ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.

In 2014, Rogan et al. \cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM, to the Long Range Energy Alternatives Planning System (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in the transport, industry and services sectors in Ireland. The bottom-up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS. Three scenarios are explored, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario. The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020. In the reference scenario, the residential total energy consumption rose 6.7 \% in 2020, compared to 2008. The share of space and water heating decreased from 82 \% to 77 \% in that period. In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% reduction of energy consumption in 2020, compared to the reference scenario. The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits. The results in a reduction in of energy consumption of 16.5 \% in 2020, compared to the reference scenario.Rogan et al. discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements in the residential sector. The importance for the policy to not only look at the number of retrofits, but also the depth of retrofit work, is shown by the difference in scenarios. The model also found considerable scope for energy savings to be made through improved building regulations.

In 2018, Thellufsen et al. \cite{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of district heating in both the commercial and residential sectors of Ireland. The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ reduction in 2050, compared to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN. The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland. A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates, however in all cases the inclusion of district heating is optimal. The final results find that district heating reduces fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal. Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.

@@

Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector. This was the first comprehensive analysis of the residential sector, and their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}. In 2000, Clinch and Healy developed and published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating. A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts. Under the predicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
The same model was used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component
into the computer model. The modelling process involved developing a computer model of the energy performance of the existing dwelling stock. It was noted that the energy consumption predicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions. 
The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.

In 2011, Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP). The model focused on the energy efficiency of space and water heating by examining technology and building fabric efficiency. The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations. Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model predicts a reduction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperatures affect on energy efficiency improvements.  
\par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock. A base geometry according to age bands and thermal characteristics was created in the study. The heat load for a statistical occupancy and year was established. The default values used in some of the BER database variables where highlighted. For example,  there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey found 18\% of detached housing in Ireland had no roof insulation. Ahern et al. found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area. Concluding that the majority of energy efficiency progress for thermal use in Irish dwellings is being offset. The model thermal improvement measures resulted in a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback. Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.

@@

The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}. ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates the annual energy consumption to simulate the potential for improved energy efficiency gains within the existing retrofit programs. Mac Uidhir et al. specifically examined the Better Energy Homes (BEH) scheme. The five most common retrofit combinations are simulated for each of the 9 archetypes ( 3 energy rating types and 3 building types ). The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when comparing the baseline scenario. This shows the the type of retrofit and type of house can have a large affect on the energy saved. 
Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is considered negligible.

@@

The models described in this section include a range of national ESOMs and bottom-up residential models. The models have varied objectives such as exploring consumer choices, PV battery system location, evaluating the health effects of retrofitting, and integration of bioenergy and heat. The thermal efficiency of dwellings is a common objective in Irish models, due to the poor thermal efficiency of the dwelling stock and the growing BER database is providing high data granularity input to these models. The internal temperature is also analysed in some of the studies, a recurring theme is the wide range of internal temperatures. 
The bottom-up residential models are typically simulation models looking at a specific policy. The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switching to Microsoft SQL. 
There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferred when focusing on that sector. The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation, or transport. The use of ESOM to explore DH in Ireland allows for insights which bottom-up simulation models cannot provide. The connection between waste heat in electricity and industry can show the optimal solution for district heating in commercial and residential sectors.

@@

Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings, and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx. 85\% of residential buildings are 2005 or older. Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil, accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.

@@

\begin{figure}[!htbp]
    \begin{tikzpicture}[every node/.style={align=center}, pin distance=17mm]
    \pie[cloud,text=inside,radius = 2.4]{34/Agriculture,7/Other,59/Energy-related}
    \draw [->, line width=4pt] (5.4,-0.3) -- (6.5,-.3);
    \hspace{8.5cm}
    \pie[rotate =30,text=inside,radius=1.95,color={black!30,red!70,yellow!40}]{40/Transport,33/Heat,27/Electricity}
    \end{tikzpicture}
    \caption{Ireland 2018 GHG Share \cite{Howley2005ENERGY-RELATEDIreland}}
    \label{fig:Ireland2018GHG}
\end{figure}

@@

Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018. The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}. These targets account for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to make sectoral five-year carbon budgets a legal requirement to aid progress of the forementioned long-term targets. 
\par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG reduction target (first NDC), has been succeeded by a new target of at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}. Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. Climate Action Plan 2021 will have to be more ambitious than its predecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national target also.
There are also renewable energy targets to consider for transport (RES-T), heat (RES-H), and electricity (RES-E) under the renewable directive (EU) 2018/2001, and the energy efficiency targets under (EU) 2018/2002.

@@

%The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to reduce Non-ETS GHG by 30\% in 2030 compared to 2005 levels. 
%The x-axis (width of each bar chart) shows the potential reduction of annual $MtCO2_{eq}$. The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_{eq}$ over the 2021 to 2030 period. The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_{eq}$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
%The options specific to the residential sector are:

@@

%\begin{figure}[!htbp]
 %\centering
 %\includegraphics[scale=1]{Figures/CAP2019MACC.png} 
 %\caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
% \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
%\end{figure}
%For example, Figure \ref{fig:CAP2019:MACC} shows the retrofitting of oil boilers is expected to reduce the Total Cost of Ownership (TCO) (ie. save money), but only abate a small amount of carbon, whereas switching oil boilers to heat pumps will increase TCO (ie. cost money) but the expected abatement of carbon is far greater.

@@

Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy. A €15/ $tCO_2$ carbon tax was introduced in 2009 to reduce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030. The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC. REFIT operates by guaranteeing a minimum price for electricity over a 15 year period. The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020. RESS supports renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target. 

@@

THe cost data of appliances 
Topten -

>>>>

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{pgf-pie}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{anyfontsize}
\usepackage{pgfplots}
\usepackage{amsmath}
\usepackage{comment}
\usepackage{booktabs}
\usepackage{rotating}
\usepackage{siunitx}
\usepackage{eurosym}
\usepackage{multirow}
\usepackage{xcolor} 
\usepackage{colortbl}

@@

Energy System Optimisation Models (ESOMs) can be used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.

\par For example, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK using UK TIMES Model (UKTM). As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. Li set out to incorporate a Discrete Choice Model (DCM) for TIMES to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices. While Li's research has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights through more surveys and analysis.

@@

Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark. The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland, and Germany to cover import/export. Exogenous user constraints applied to the model, based on the current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030, and individual oil boilers are phased out by 2030. The results find that when investments in individual heat pumps are allowed, air-to-water heat pumps are installed in non-district heating areas. This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and orangeucing system costs, fuel consumption, and $CO_2$ emissions.
Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a orangeuced need for peak/reserve capacity investments. The Balmorel model has provided insights, but it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies. Nevertheless, the insights are valuable and the results can be soft-linked to other models for deeper analysis.

Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at the household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island. 
Two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach. Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be stoorange in communal batteries. When identical communal and individual battery capacities where compaorange, it was fond  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery. While the battery location was exploorange, the PV system location was not, further investigation into the optimal PV system location could provide further insights. The household occupancy behaviour with and without household batteries on the PV system could alter the optimal pathway.

@@

ESOM models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem,Connolly2021AalborgEnergy-System}, however few Irish studies use ESOM models to explore the residential sector, two examples of Irish ESOM focusing on the residential sector are detailed here.

In 2014, Rogan et al. \cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM, to the Long Range Energy Alternatives Planning System (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in the transport, industry and services sectors in Ireland. The bottom-up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS. Three scenarios are exploorange, namely, the reference scenario, energy efficiency scenario, and energy efficiency+ scenario. The reference scenario assumes that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020. In the reference scenario, the residential total energy consumption rose 6.7 \% in 2020, compaorange to 2008. The share of space and water heating decreased from 82 \% to 77 \% in that period. In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% orangeuction of energy consumption in 2020, compaorange to the reference scenario. The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits. The results in a orangeuction in of energy consumption of 16.5 \% in 2020, compaorange to the reference scenario.Rogan et al. discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements in the residential sector. The importance for the policy to not only look at the number of retrofits, but also the depth of retrofit work, is shown by the difference in scenarios. The model also found considerable scope for energy savings to be made through improved building regulations.

In 2018, Thellufsen et al. \cite{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of district heating in both the commercial and residential sectors of Ireland. The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ orangeuction in 2050, compaorange to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN. The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland. A sensitivity analysis was performed on investment costs, heat grid losses, and discount rates, however in all cases the inclusion of district heating is optimal. The final results find that district heating orangeuces fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal. Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.

@@

Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector. This was the first comprehensive analysis of the residential sector, and their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}. In 2000, Clinch and Healy developed and published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating. A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts. Under the porangeicted scenario, energy savings easily outweighed programme costs, representing a benefit-cost ratio of 1.7.The benefits of orangeuced sickness, hospitalisation, drugs, and mortality were quantified and this moved the  benefit-cost ratio up to 3.0.
The same model was used by Clinch, Healy, and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component
into the computer model. The modelling process involved developing a computer model of the energy performance of the existing dwelling stock. It was noted that the energy consumption porangeicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions. 
The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compaorange to the reference scenarios.

In 2011, Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP). The model focused on the energy efficiency of space and water heating by examining technology and building fabric efficiency. The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations. Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model porangeicts a orangeuction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperatures affect on energy efficiency improvements.  
\par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock. A base geometry according to age bands and thermal characteristics was created in the study. The heat load for a statistical occupancy and year was established. The default values used in some of the BER database variables where highlighted. For example,  there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey found 18\% of detached housing in Ireland had no roof insulation. Ahern et al. found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area. Concluding that the majority of energy efficiency progress for thermal use in Irish dwellings is being offset. The model thermal improvement measures resulted in a orangeuction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback. Houses built in 1979 or after resulted in an average 26\% orangeuction in running costs and $CO_2$ emissions from retrofit measures.

@@

The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}. ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates the annual energy consumption to simulate the potential for improved energy efficiency gains within the existing retrofit programs. Mac Uidhir et al. specifically examined the Better Energy Homes (BEH) scheme. The five most common retrofit combinations are simulated for each of the 9 archetypes ( 3 energy rating types and 3 building types ). The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when comparing the baseline scenario. This shows the the type of retrofit and type of house can have a large affect on the energy saved. 
Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is consideorange negligible.

@@

The models described in this section include a range of national ESOMs and bottom-up residential models. The models have varied objectives such as exploring consumer choices, PV battery system location, evaluating the health effects of retrofitting, and integration of bioenergy and heat. The thermal efficiency of dwellings is a common objective in Irish models, due to the poor thermal efficiency of the dwelling stock and the growing BER database is providing high data granularity input to these models. The internal temperature is also analysed in some of the studies, a recurring theme is the wide range of internal temperatures. 
The bottom-up residential models are typically simulation models looking at a specific policy. The complexity of models has increased significantly in the last 20 years, in Ireland, the models started out in Microsoft Excel and switching to Microsoft SQL. 
There has been few studies on the Irish residential sector using ESOMs, as residential bottom-up models are preferorange when focusing on that sector. The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation, or transport. The use of ESOM to explore DH in Ireland allows for insights which bottom-up simulation models cannot provide. The connection between waste heat in electricity and industry can show the optimal solution for district heating in commercial and residential sectors.

@@

Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings, and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% orangeuction of CO\textsubscript{2} emissions in new dwellings compaorange to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx. 85\% of residential buildings are 2005 or older. Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil, accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.

@@

\begin{figure}[!htbp]
    \begin{tikzpicture}[every node/.style={align=center}, pin distance=17mm]
    \pie[cloud,text=inside,radius = 2.4]{34/Agriculture,7/Other,59/Energy-related}
    \draw [->, line width=4pt] (5.4,-0.3) -- (6.5,-.3);
    \hspace{8.5cm}
    \pie[rotate =30,text=inside,radius=1.95,color={black!30,orange!70,yellow!40}]{40/Transport,33/Heat,27/Electricity}
    \end{tikzpicture}
    \caption{Ireland 2018 GHG Share \cite{Howley2005ENERGY-RELATEDIreland}}
    \label{fig:Ireland2018GHG}
\end{figure}

@@

Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to orangeuce Greenhouse Gas (GHG) emissions by 51\% in 2030 compaorange to 2018. The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}. These targets account for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to make sectoral five-year carbon budgets a legal requirement to aid progress of the forementioned long-term targets. 
\par The EU's first Nationally Determined Contribution (NDC), which had a 40\% GHG orangeuction target (first NDC), has been succeeded by a new target of at least a 55\% GHG orangeuction target (updated NDC) by 2030 compaorange to 1990 levels \cite{SUBMISSIONSTATES}. Ireland's Climate Action Plan 2019 (CAP2019) complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG orangeuction target in 2030 compaorange to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. Climate Action Plan 2021 will have to be more ambitious than its porangeecessor, CAP2019 as it must reflect EU's updated NDC target and take account of the national target also.
There are also renewable energy targets to consider for transport (RES-T), heat (RES-H), and electricity (RES-E) under the renewable directive (EU) 2018/2001, and the energy efficiency targets under (EU) 2018/2002.

@@

%The CAP2019 MACC, shown in Fig.\ref{fig:CAP2019:MACC} provides $>300$ of the most cost-effective technologies options to orangeuce Non-ETS GHG by 30\% in 2030 compaorange to 2005 levels. 
%The x-axis (width of each bar chart) shows the potential orangeuction of annual $MtCO2_{eq}$. The y-axis (height of each bar chart ) shows the associated average cost of abating one tonne of $CO2_{eq}$ over the 2021 to 2030 period. The columns are organised from the most economical (left side) to the most expensive technology (right side) in $EUR/tCO2_{eq}$ \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}.
%The options specific to the residential sector are:

@@

%\begin{figure}[!htbp]
 %\centering
 %\includegraphics[scale=1]{Figures/CAP2019MACC.png} 
 %\caption{Climate Action Plan 2019 - Marginal Abatement Cost Curve}
% \label{fig:CAP2019:MACC} \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}
%\end{figure}
%For example, Figure \ref{fig:CAP2019:MACC} shows the retrofitting of oil boilers is expected to orangeuce the Total Cost of Ownership (TCO) (ie. save money), but only abate a small amount of carbon, whereas switching oil boilers to heat pumps will increase TCO (ie. cost money) but the expected abatement of carbon is far greater.

@@

Ireland uses both the \textit{carrot} (subsidy) and \textit{stick} (tax) approach to direct private investment towards low carbon or renewable energy. A €15/ $tCO_2$ carbon tax was introduced in 2009 to orangeuce investment in fossil fuels, and the increasing carbon tax rate has resulted in tax receipts growing every year from 2010 to 2018 \cite{Oireachtas2019AnOffice}. Ireland's current carbon tax of €33.50/ $tCO_2$ will rise to €100/ $tCO_2$ in 2030. The Renewable Energy Feed-in Tariff (REFIT) schemes were designed to incentivise renewable electricity investment to achieve legally binding renewable targets under 2009/28/EC. REFIT operates by guaranteeing a minimum price for electricity over a 15 year period. The Renewable Electricity Support Scheme (RESS) replaced REFIT in 2020. RESS supports renewable electricity projects to help Ireland achieve 2030 climate targets, especially the 70\% renewable electricity target. 

@@

\par The cost data for different electrical appliances was sourced from Topten EU database. Topten has a large database covering electrical appliances available in the European market, the data collected includes cost and efficiency. While some Topten reports have looked at sales and market shares \cite{Michel2016EnergyREPORT}.

\subsubsection{Savings}
\label{LitR:Savings}
